Discovering governing reactions from concentration data

### 2018-02-01 docs/manuscript.tex
manuscript update case 1

	When presented with a time series consisting of $T$ observations, the data can be represented as a matrix
	\begin{align}
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\end{align}
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\begin{align}
		\theta_r(\textbf{X}) = \begin{pmatrix}
		\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
		\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\label{method:the-reactions}\end{align}
	where $\textbf{X}_i$ denotes the $i$-th row in $X$. Applying the concentration trajectory to the library yields $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\begin{align}
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\end{align}
	In particular, the system is linear in the coefficients $\Xi$, which makes potentially sparse regression tools such as elastic net regularization \cite{Zou2005} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\lambda\|\Xi\|_1 + \alpha(1-\lambda)\|\Xi\|_2^2 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\|\cdot\|_F$ denotes the Frobenius norm, $\lambda\in[0,1]$ a hyperparameter that interpolates linearly between LASSO \cite{Tibshirani1996, Hastie2009} and Ridge \cite{Hoerl1} methods, and $\alpha\geq 0$ is a hyperparameter that, depending on $\lambda$, can induce sparsity and give preference to smaller solutions in the $L_1$ or $L_2$ sense.

	When presented with multiple trajectories one can extend the objective function by a sum over these and obtain a more general minimization problem that reads: Find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \sum_{\text{traj}}\frac{1}{2T_{\text{traj}}}\left\| \dot{\textbf{X}}_\text{traj} - \Theta(\textbf{X}_\text{traj})\Xi \right\|_F^2 + \alpha\lambda\|\Xi\|_1 + \alpha(1-\lambda)\|\Xi\|_2^2 \right)
	\end{align}
	subject to $\Xi\geq 0$.
	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) the sequential least-squares minimizer SLSQP \cite{Kraft1988} is applied via the software package SciPy \cite{SciPy}. Since only the concentration data $\mathbf{X}$ is available but not its temporal derivative $\dot{\mathbf{X}}$, it is approximated numerically by second order finite differences with the exception of boundary data.

@@

    \section{Scenarios of measurement}\label{sec:scenarios-of-measurement}

    The proposed estimation method is designed to find reaction networks and rates from experimental data. In the following measurement scenarios are considered that differ in how many time series are available and the level of noise that the data is subject to:
    \begin{enumerate}
        \item The initial state of the measurement apparatus can be reproduced exactly. There is only one set of generating reaction rates. Each frame of a time series is subject to noise. The outcome of this scenario is one time series of concentrations with a certain noise-level.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. The outcome of this scenario is multiple time series each with a certain noise level and different initial states.
        \item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. Additionally the generating reaction rates differ slightly in each observation. The outcome of this scenario is multiple time series each with a given noise level, different initial states and slightly perturbed underlying reaction rates.
    \end{enumerate}

	\section{Examples}\label{sec:examples}

	In the following two examples are presented. The first example considers data according to the first scenario of measurement as presented in Section \ref{sec:scenarios-of-measurement}.
	The minimization problem (\ref{method:minimizationproblem}) with $\alpha=0$ is solved for different magnitudes of noise variance.

	The second example deals with the second scenario of measurement and solves the minimization problem (\ref{method:minimizationproblem}) with regularization parameters being determined by cross validation.

	In each of the considered examples the library $\Theta$ includes, among others, the reactions of the example system described in Section \ref{sec:generegulation}. The full set of reactions can be found in the Appendix, Table \ref{tab:reaction-library}.

	\subsection{Regression without regularization}
	\label{sec:ex1}

	In this example it is demonstrated that given enough data, the original model can be recovered in the limit of the noise variance $\sigma^2\to 0$. To this end, we first solve

	\begin{colbox}{F8E0E0}
		todo
	\end{colbox}

	Here we demonstrate that, with decreasing time step $\Delta t$ and $\alpha=0$, the estimated rates converge to the true rates if there is no ambiguity in the system. Let there be two species $A$ and $B$ which can unimolecularly convert into each other with macroscopic rates of $k_1=4.0$ and $k_2=0.5$, respectively. The initial amount of particles is set to $70$ of type $A$ and $0$ of type $B$. The law of mass action solution and Gillespie realization corresponding to this system are depicted in Figure \ref{ex1:concentration-curves}.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_ex1_concentration_curves}
		\end{center}
		\caption{The law of mass action solution and a Gillespie realization of the example given in Section \ref{sec:ex1}.}
		\label{ex1:concentration-curves}
	\end{figure}

	For each $\Delta t$, eight different realizations with identical starting conditions are averaged into a single trajectory, of which then the rates for reactions $A\to B$ and $B\to A$ are estimated. This procedure was repeated $30$ times. In Figure \ref{ex1:rates-convergence}, the average estimates with their standard deviations are shown. One can see that, with decreasing $\Delta t$, the estimates become better and the rates can be recovered. Further it can be observed that in this case a time step of $\Delta t\approx 10^{-2}$ suffices.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_rates_convergence}
		\end{center}
		\caption{Convergence of estimated rates with decreasing $\Delta t$.}
		\label{ex1:rates-convergence}
	\end{figure}

	\subsection{Regression with regularization}
	\label{sec:ex2}
	In this example there are four species $A$, $B$, $C$, and $D$. The system was generated with reactions $A\xrightarrow{k_1}D$, $D\xrightarrow{k_2}A$, $D\xrightarrow{k_3}B$, $A+B\xrightarrow{k_4}C$, and $C\xrightarrow{k_5}D+B$, where the rates are given in Table \ref{tab:estimated-rates}. These reactions are inserted into the ansatz library $\Theta (\cdot)$. Additionally the reactions $A\leftrightarrow B$, $A\leftrightarrow C$, and $A+C\leftrightarrow D$ are inserted into the library.

	\begin{figure}
		\centering
		\begin{subfigure}[b]{.45\textwidth}
			\includegraphics[width=\textwidth]{./figures_tex/cv_concentrations_train}
			\caption{Train data set.}
		\end{subfigure}
		~
		\begin{subfigure}[b]{.45\textwidth}
			\includegraphics[width=\textwidth]{./figures_tex/cv_concentrations_test}
			\caption{Test data set.}
		\end{subfigure}
		\caption{Train and test data sets for the example given in Section \ref{sec:ex2}.}
		\label{ex2:test-train-concentration-curves}
	\end{figure}

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\dot{\mathbf{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score is $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/cv_score}
		\end{center}
		\caption{Score function (\ref{ex2:score-function}) evaluated for a range of $\alpha$ values.}
		\label{ex2:score-graph}
	\end{figure}

	One can see that the minimum score is achieved at
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with $\hat{\alpha}$ as hyperparameter for the regularizer and applying a cut-off of $10^{-10}$ to the yielded parameters, one obtains the simpler set of reactions $A\xrightarrow{\hat{k}_1}D$, $D\xrightarrow{\hat{k}_3} B$, and $A+B\xrightarrow{\hat{k}_4}C$ with rates as in Table \ref{tab:estimated-rates}.

	\begin{table}
		\centering
		\begin{tabular}{ccc}
			Reaction & original rate & estimated rate \\ \hline\noalign{\vskip .1cm}
			$A \to D$ & $k_1 = 2\cdot 10^{-2}$ & $\hat{k}_1 \approx 3.5\cdot 10^{-3}$\\
			$D \to A$ & $k_2 = 2\cdot 10^{-2}$ & $\hat{k}_2 = 0$ \\
			$D\to B$ & $k_3 = 1\cdot 10^{-2}$ & $\hat{k}_3\approx 9.1 \cdot 10^{-3}$ \\
			$A+B\to C$ & $k_4 = 1\cdot 10^{-4}$ & $\hat{k}_4\approx 8.7 \cdot 10^{-5}$ \\
			$C \to D+B$ & $k_5 = 1\cdot 10^{-4}$ & $\hat{k}_5 = 0$
		\end{tabular}
		\caption{Original and estimated reaction rates for the example in Section \ref{sec:ex2}.}
		\label{tab:estimated-rates}
	\end{table}

	One can see in Figure \ref{ex2:result} that with this simpler set of reactions the observed concentration curves can be explained as well. Further, it can be argued that the second reaction of the original model $D\xrightarrow{k_2}A$ is compensated by a smaller rate $\hat{k}_1 < k_1$ and that the fifth reaction of the generating model can be neglected altogether in the observed timescale as its rate is two orders of magnitude smaller than the other unimolecular reactions' rates.
	The parameters $k_3$ and $k_4$ are approximately recovered by $\hat{k}_3$ and $\hat{k}_4$, respectively.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/cv_expected_estimated_concentrations}
		\end{center}
		\caption{Integrated law of mass action equations (\ref{method:the-system}) for the sparse estimated parameters and the corresponding Gillespie realization for the example in Section \ref{sec:ex2}.}
		\label{ex2:result}
	\end{figure}

	\section{Conclusion}
	In this work we have successfully applied and extended the SINDy method to not only parsimoniously detect potentially nonlinear terms in a dynamical system from noisy data, but also yield, in this case, a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).

@@

	\newpage
% 	\bibliographystyle{alpha}
	\bibliographystyle{abbrv}
	\bibliography{bibliography.bib}

	\section{Appendix}
	\begin{table}

@@

@@

>>>>

	When presented with a time series consisting of $T$ observations, the data can be represented as a matrix
	\begin{align}
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\end{align}
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\begin{align}
		\theta_r(\textbf{X}) = \begin{pmatrix}
		\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
		\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\label{method:the-reactions}\end{align}
	where $\textbf{X}_i$ denotes the $i$-th row in $\textbf{X}$. Applying the concentration trajectory to the library yields $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\begin{align}
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\end{align}
	In particular, the system is linear in the coefficients $\Xi$, which makes potentially sparse regression tools such as elastic net regularization \cite{Zou2005} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\lambda\|\Xi\|_1 + \alpha(1-\lambda)\|\Xi\|_2^2 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\|\cdot\|_F$ denotes the Frobenius norm, $\lambda\in[0,1]$ a hyperparameter that interpolates linearly between LASSO \cite{Tibshirani1996, Hastie2009} and Ridge \cite{Hoerl1} methods, and $\alpha\geq 0$ is a hyperparameter that, depending on $\lambda$, can induce sparsity and give preference to smaller solutions in the $L_1$ or $L_2$ sense.

	When presented with multiple trajectories the data matrices $\mathbf{X}$ and $\dot{\mathbf{X}}$ can be stacked and inserted into the minimization problem.

	For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) the sequential least-squares minimizer SLSQP \cite{Kraft1988} is applied via the software package SciPy \cite{SciPy}. Since only the concentration data $\mathbf{X}$ is available but not its temporal derivative $\dot{\mathbf{X}}$, it is approximated numerically by second order finite differences with the exception of boundary data.

@@

	\section{Proof of concept}\label{sec:proof-of-concept}

	The proposed estimation method is designed to find reaction networks and rates from experimental data. In the following the method is applied to data generated from the example described in Section \ref{sec:generegulation}.

	First it is assumed that the concentration data can be obtained without noise and it is shown that while the minimization problem (\ref{method:minimizationproblem}) without regularization, i.e., $\alpha=0$, yields rates that fit well to the input data, the original rates cannot be recovered while adding $L_1$ regularization improves the situation and the original model \textcolor{red}{almost} can be recovered.

	In Section \ref{sec:case-2} the example data contains additive Gaussian noise. It is assumed that the initial state can be reproduced exactly and one has several measurements. It is shown that in the limit of small noise one can obtain a good fit and regularization helps in the reliability of the estimated reaction rates.

	The last example is described in Section \ref{sec:case-3} and can be seen as an extension of the situation in Section \ref{sec:case-3} to multiple initial conditions.

	%measurement scenarios are considered that differ in how many time series are available and the level of noise that the data is subject to:
	%\begin{enumerate}
	%	\item The initial state of the measurement apparatus can be reproduced exactly. There is only one set of generating reaction rates. Each frame of a time series is subject to noise. The outcome of this scenario is one time series of concentrations with a certain noise-level.
	%	\item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. The outcome of this scenario is multiple time series each with a certain noise level and different initial states.
	%\item The process is observed multiple times. The initial state of the measurement apparatus differs in each observation. Additionally the generating reaction rates differ slightly in each observation. The outcome of this scenario is multiple time series each with a given noise level, different initial states and slightly perturbed underlying reaction rates.
	%\end{enumerate}

	\subsection{The noiseless case}\label{sec:case-1}

	In this section noiseless data generated from the law of mass action system described in Section \ref{sec:generegulation} is applied to the minimization problem (\ref{method:minimizationproblem}). When setting the hyperparameter $\alpha=0$ one obtains least squares regression with a constraint on the non-negativity of the estimated reaction rates.

	\begin{figure}
		\centering
		\includegraphics[width=.7\textwidth]{./figures_tex/bubbles}
		\caption{Generating and estimated reaction rates in the situation of Section \ref{sec:case-1}. The y and x axes contain reaction educts and products, respectively. The black squares denote the reactions with which the system was generated and their size indicates the respective reaction rate. The blue circles are least squares estimated reaction rates and the red rings depict rates which were obtained by some $\alpha>0$ in the minimization problem (\ref{method:minimizationproblem}).}
		\label{fig:case-1-sparsity-pattern}
	\end{figure}

	When applying no regularization, one can observe in Figure \ref{fig:case-1-sparsity-pattern} that the sparsity pattern does not match the generating reaction network and the reaction responsible for the decay of $\mathrm{A}$ particles is completely ignored. For a suitable choice of the hyperparameter $\alpha > 0$ one can obtain a sparser solution and also recover the decay reaction as indicated by the red rings in the figure.

	\subsection{Multiple measurements at the same noise level}\label{sec:case-2}

	\subsection{Multiple initial conditions}\label{sec:case-3}

	\section{Conclusion}
	\begin{colbox}{F8E0E0}
		todo
	\end{colbox}
	In this work we have successfully applied and extended the SINDy method to not only parsimoniously detect potentially nonlinear terms in a dynamical system from noisy data, but also yield, in this case, a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).

@@

	\newpage
% 	\bibliographystyle{alpha}
	\bibliographystyle{abbrv}
	\bibliography{bibliography.bib}
	\newpage	
	\section{Appendix}\label{sec:appendix}
	\begin{table}[h]

@@

		\scalebox{0.7}{

@@

		}
